---
title: "USA SHOOTINGS"
output: html_notebook
---
Details about this project can be found in the readme file 
```{r}
getwd()
```

#### **import the required libraries** 
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(lubridate)
library(lattice)

print("Setup Complete!")
```

#### **Import the Data-set**
```{r}
filepath <- "F:/Data Science/My Datasets/USA_shootings.csv"
shootings <- read.csv(filepath)
print("Import Successful")
```

# **EXPLORATORY DATA ANALYSIS**
```{r}
head(shootings)
```

``````{r}
dim(shootings)
```

* There are 4895 victims and 10 observations surrounding their death

#### **Addition of the Age group column** 
```{r}
shootings <- shootings%>% mutate(Age_group = case_when(age > 0 & age <= 12 ~ "Child",
                                                                 age >= 13 & age <= 19 ~ "Teenager",
                                                                 age >= 20 & age <= 40 ~ "Young Adult",
                                                                 age >= 41 & age <= 65 ~ "Adult",
                                                                 age > 65 ~ "Senior"))
```

Now lets check if the column has been added 
```{r}
head(shootings)
```
Tadaaaaa>>>>

#### **Conversion of columns to appropriate classes**
```{r}
shootings$manner_of_death <- as.factor(shootings$manner_of_death)
shootings$armed <- as.factor(shootings$armed)
shootings$date <- ymd(shootings$date)
shootings$gender <- as.factor(shootings$gender)
shootings$race <- as.factor(shootings$race)
shootings$city <- as.factor(shootings$city)
shootings$state <- as.factor(shootings$state)
shootings$signs_of_mental_illness <- as.factor(shootings$signs_of_mental_illness)
shootings$threat_level <- as.factor(shootings$threat_level)
shootings$flee <- as.factor(shootings$flee)
shootings$body_camera <- as.factor(shootings$body_camera)
shootings$arms_category <- as.factor(shootings$arms_category)
```

Lets check if the classes of the columns have been changed 
```{r}
head(shootings)
```
## **Manner Of Death**
```{r}
mann.tbl <- shootings %>% group_by(manner_of_death) %>% summarize(`Death Toll` = n())
mann.tbl <- rename(mann.tbl, "Manner of Death" = "manner_of_death")
mann.tbl
```
* As seen above, only few of the victims were shot and tasered 

```{r}
g <- ggplot(mann.tbl, aes(x = `Manner of Death`, y = `Death Toll`))
g + geom_bar(stat = "identity") + ggtitle("DEATH TOLL BY MANNER OF DEATH") + theme_classic()
```
* The chart above clearly shows the wide difference between individuals that were only shot and individuals that were shot and tasered 


## **Armed** 
```{r}
armed_grp <- table(shootings$armed)
armed_grp <- as.data.frame(armed_grp)
armed_grp <- rename(armed_grp, "Weapons" = "Var1", "Victims Count" = "Freq")
#new_column_name = old_column_name
armed_grp <- armed_grp %>% arrange(desc(`Victims Count`)) %>% mutate(Percentage = (`Victims Count`/4895) * 100)
armed_grp$Percentage <- round(armed_grp$Percentage, digits = 1)
armed_grp
```
* As seen above most of the victims were armed with Guns (56%) and Knives (14.1%). The victims might have tried to shoot at or stab the police officer before being shot dead

Lets move further to uncover more insights 

```{r}
guns <- subset(shootings, armed == "gun")
gun.grp <- guns%>% select(race) %>% group_by(race) %>% summarize(`Owned Gun` = n()) %>% arrange(desc(`Owned Gun`))
print(gun.grp)
```
* More white victims owned guns more than any other race in this data set

## **Gender and Race**
```{r}
gr_grp <- shootings %>% select(race, gender) %>% group_by(race, gender) %>% summarize(`Victim Counts` = n())

gr_grp
```

```{r}
g <- ggplot(gr_grp, aes(x = race, y = `Victim Counts`))
g + geom_bar(aes(fill = gender), stat = "identity", position = position_dodge()) + xlab("RACE") + ylab("DEATH TOLL") + ggtitle("DEATH TOLL BY RACE AND GENDER") + scale_fill_manual(values = c("green", "navy blue")) + theme_light()
```

* It ia evident that police shot more white men and women than any other race as opposed to popular belief that blacks have been the more victims.

Lets dive deeper 

## **City**
```{r}
city.grp <- shootings %>% select(city) %>% group_by(city) %>% summarize(`Death Toll` = n()) %>% arrange(desc(`Death Toll`)) %>% mutate(Percentage = (`Death Toll`/4895) *100)

city.grp$Percentage <- round(city.grp$Percentage, digits = 1)
city.grp
```
* The top three cities where victims were killed are Los Angeles, Phoenix and Houston. There are 2288 cities in the data-set 

## **State**
```{r}
states_grp <- shootings %>% select(state) %>% group_by(state) %>% summarize(`Death Toll` = n()) %>% arrange(desc(`Death Toll`)) %>% mutate(Percentage = (`Death Toll`/4895) * 100)
states_grp$Percentage <- round(states_grp$Percentage, digits = 1)

states_grp
```
*  As seen above the top three states with the highest Death toll are California, Texas and Florida 

## **Mental Status**
```{r}
mental.illness <-table(shootings$signs_of_mental_illness)
mental.illness
```

* Most of the victims didn't show signs mental illnesses 

```{r}
ment <- data.frame(`Mental status` = c("No mental Illness", "Has mental illness"), `Death toll` = c(3792, 1103))
ment
```

```{r}
g <- ggplot(ment, aes(x = "", y = Death.toll, fill = Mental.status))
bar <- g + geom_bar(stat = "identity")
bar + coord_polar("y", start = 0) + scale_fill_manual(values = c("red", "green")) +ggtitle("CHART SHOWING THE DEATH TOLL BY MENTAL STATUS \n OF THE VICTIMS")
```

## **Threat level**
```{r}
threat <- shootings %>% select(threat_level, race) %>%group_by(race,threat_level) %>%  summarize(Death.toll = n()) %>% arrange(desc(Death.toll)) 
threat <- rename(threat, "Threat level" = "threat_level")
threat
```
* Majority of the victims were a potential threat to the police hence they were shot

Lets see how the death toll by threat level regarding race in a chart below 
```{r}
g <- ggplot(threat, aes(x = race, y = Death.toll, fill = `Threat level`))
g + geom_bar(stat = "identity", position = position_dodge()) + scale_fill_manual(values = c("navy blue", "green", "red")) + ggtitle("DEATH TOLL BY RACE AND THREAT LEVEL") + theme_bw()
```
* The white race leads in all categories 

## **Was the Victim Fleeing**
```{r}
flee <- shootings %>% select(flee) %>% group_by(flee) %>% summarize(`Death toll` = n()) %>% arrange(desc(`Death toll`))
flee <- rename(flee, "Fleeing status" = "flee")
flee
```
* Most of the victims were not fleeing 

## **Was the incident recorded on body camera**
```{r}
camera <- shootings %>% select(body_camera) %>% group_by(body_camera) %>% summarize(`Death toll` = n()) %>% arrange(desc(`Death toll`))
camera
```

* Most of the incidents were not recorded on body camera 

## **Arms Category**
```{r}
arms <- shootings %>% select(arms_category, armed) %>% group_by(arms_category, armed) %>% summarize(`Death toll` = n()) %>% arrange(desc(`Death toll`))
arms <- rename(arms, "Weapon"= "armed")
arms
```


```{r}
age_group <- shootings %>% select(Age_group, race, gender) %>% group_by(Age_group, gender, race) %>% summarize(`Death toll` = n())
age_group
```

```{r}
g <- ggplot(age_group, aes(x = gender, y = `Death toll`, fill = Age_group))
g + geom_bar(stat = "identity", position = position_stack()) + facet_grid(. ~ race)
```
* As seen above, most of the victims were male young adults between the ages of 18 and 35 followed by male adults between the ages of 36 and 65
